---
title: "Tidying Data Assignment"
author: ''
date: ''
output:
  html_document:
    df_print: paged
---
# Please complete all the tasks listed below. After completing the tasks download the .Rmd file and upload in the peer review item for grading.

# Additionally please write text between the code chunks explaining what each code chunk is about.

# Refer the linked online textbook in case of any issues.

Load the tidyverse library
```{r}
library(tidyverse)
```

##### Question 1. 

The built in billboard dataset is not tidy. Describe why it is not tidy and then tidy the dataset.

The built-in billboard dataset in R is not tidy due to several violations of tidy data principles. Firstly, each variable is not forming a column; instead, variables like artist, track, date.entered, wk1 (week 1), wk2 (week 2), etc., are spread across multiple columns. Secondly, each observation does not form a row; while each row represents a single track, the weekly rankings (wk1, wk2, etc.) are horizontally spread across columns rather than being stacked vertically. Lastly, the dataset combines information about tracks with metadata about the artist and the date of entry into the Billboard charts, thus failing to separate different observational units into distinct tables, complicating data analysis tasks.

```{r question-1-response}


# Tidy the dataset
tidy_billboard_long <- billboard %>%
  # First gather up all the week entries into a row for each week for each song (where there is an entry)
  pivot_longer(cols = starts_with("wk"), 
               names_to = "week", 
               values_to = "rank", 
               names_prefix = "wk") %>%
  drop_na(rank)
  #tidy_billboard_long
  
  # Then, convert the week variable to a number and figure out the date corresponding to each week on the chart
tidy_billboard_convert <- tidy_billboard_long %>%
  mutate(
    week = as.numeric(str_extract(week, "\\d+")),  # Extract numeric part from the 'week' column and convert to numeric
    date.entered = as.Date(date.entered, format = "%Y-%m-%d"),  # Make sure date.entered is in Date format
    date = date.entered + 7*(week-1)  # Calculate current date by adding weeks to date.entered
  )
  #tidy_billboard_convert

  # Sort the data by artist, track and week.
  tidy_billboard <- tidy_billboard_convert %>% arrange(artist, track, week) 

# Display the first few entries
tidy_billboard


#Here are what your first entries should be (formatting can be     different):
#>  A tibble: 5,307 x 5
#   artist  track                   date.entered  week  rank   date
 #    <chr>   <chr>                   <date>       <int> <dbl>   <date>
 #  1 2 Pac   Baby Don't Cry (Keep... 2000-02-26       1    87 2000-02-26
 #  2 2 Pac   Baby Don't Cry (Keep... 2000-02-26       2    82 2000-03-04
 #  3 2 Pac   Baby Don't Cry (Keep... 2000-02-26       3    72 2000-03-11
 #  4 2 Pac   Baby Don't Cry (Keep... 2000-02-26       4    77 2000-03-18
 #  5 2 Pac   Baby Don't Cry (Keep... 2000-02-26       5    87 2000-03-25
 #  6 2 Pac   Baby Don't Cry (Keep... 2000-02-26       6    94 2000-04-01
 #  7 2 Pac   Baby Don't Cry (Keep... 2000-02-26       7    99 2000-04-08
 #  8 2Ge+her The Hardest Part Of ... 2000-09-02       1    91 2000-09-02
 #  9 2Ge+her The Hardest Part Of ... 2000-09-02       2    87 2000-09-09
 # 10 2Ge+her The Hardest Part Of ... 2000-09-02       3    92 2000-09-16
 # … with 5,297 more rows

```

##### Question 2. 

Tidy the "fish_encounters" dataset of fish spotting by monitoring stations. Make the NA into 0 using the option "values_fill = list(seen = 0)"


```{r question-2-response, error=TRUE}


# Tidying the dataset
tidy_fish <- fish_encounters %>%
  pivot_wider(
    names_from = station,        # Define how to create new column names
    values_from = seen,          # Define which values to spread across the new columns
    values_fill = list(seen = 0) # Fill NA values with 0
  )

# View the tidied dataset
print(tidy_fish)


```

##### Question 3. 

Import the flowers1 dataset. Tidy and pivot the data. Hint: use "read_csv2()" to read in the dataset 

```{r question-3-response, warning=FALSE}

# Read the dataset
flowers_data <- read_csv2("flowers1.csv")

# Tidy the data: handle numeric conversion
flowers_data <- flowers_data %>%
  mutate(Value = as.numeric(gsub(",", ".", Value)))  # Replace commas with dots and convert to numeric

# Pivot the data to wide format
tidy_flowers <- flowers_data %>%
  pivot_wider(
    names_from = Variable,
    values_from = Value
  )

# View the tidied data
tidy_flowers

```

##### Question 4.

Import the flowers2 dataset. Tidy the dataset by turning the one column into 3 separate columns

```{r question-4-response}

# Read the dataset
flowers_data <- read_csv2("flowers2.csv", col_names = c("Flowers/Intensity", "Time"))

# Split the 'Flowers/Intensity' column into 'Flowers' and 'Intensity'
flowers_data <- flowers_data %>%
  separate(`Flowers/Intensity`, into = c("Flowers", "Intensity"), sep = "/", convert = TRUE)

# View the tidied data
flowers_data

```

##### Question 5.

In the following dataset, turn the implicit missing values to explicit 

```{r question-5-response, warning = FALSE}
output <- tibble(
      treatment   = c("a", "b", "a", "c", "b"),
      gender   = factor(c("M", "F", "F", "M", "M"), levels = c("M", "F", "O")),
      return = c(1.5, 0.75,  0.5,  1.8,  NA)
    )

# Explicitly handle missing values
output <- output %>%
  mutate(across(where(is.factor), fct_explicit_na, na_level = "Missing"),
         across(where(is.numeric), ~ replace_na(., 0)))  # Replacing NA in numeric columns with 0 (if needed)

# Print the modified tibble
output

```


#####Question 6.

Import the weather dataset as weather. Use "pivot_longer()" to put the days all in one column, then use "pivot_wider" to separate tmax and tmin into separate columns. Print the summary of the final resulting dataset 

```{r question-6-response}

# Read the dataset
weather <- read_csv("weather.csv")

# Pivot to longer format
weather_long <- weather %>%
  pivot_longer(
    cols = starts_with("d"),
    names_to = "day",
    values_to = "value"
  )

# Pivot to wider format
weather_wide <- weather_long %>%
  pivot_wider(
    names_from = element,
    values_from = value
  )

#Final Data Set
weather_wide

# Print the summary of the final dataset
summary(weather_wide)
  

```

###### Question 7. 

Load the built in "anscombe" data frame and use "pivot_longer()" to separate all the x and y columns and categorize them into 4 sets

```{r question-7-response}

# Reshape the dataset into a longer format
anscombe_long <- anscombe %>%
  pivot_longer(
    everything(), 
    names_to = c(".value", "set"), 
    names_pattern = "(.)(.)"
  )

# Print the transformed data
anscombe_long

```
